---
title: "STAT 431 Project 1 Final Deliverable"
author: "Julia Aptekar"
embedded-resources: true
code-fold: true
format: html
---
```{r}
#| echo: false
#install.packages("tidyverse")
#install.packages("wesanderson")
#install.packages("gt")
#install.packages("broom")
options(pillar.width = Inf)
library(tidyverse)
library(wesanderson)
library(gt)
library(broom)
democracy_data <- read.csv("Combined_Democracy_Data.csv")
```

## Explore the data

Our study utilized two datasets. The first was compiled by Christian Bjørnskov and Martin Rode, two economists who study political institutions and how governments change over time. Their goal was to create a resource for researchers studying political transitions, an area they felt had not been adequately covered by existing data. The dataset collects information on 192 sovereign countries and 16 self-governing territories annually from 1950 to 2020 on each country's regime type, democratic status, elections, leadership, suffrage, transfers of power, and colonial history. In total, it includes 14,768 country-year observations across 43 variables.

To build this dataset, the authors expanded on an existing collection of political data by recording key features of each country's government, such as whether elections took place, whether those elections were considered free and fair, and whether leadership changed hands peacefully. Records of coups were verified across newspaper archives, encyclopedias, and other databases to ensure accuracy. For our analysis, this dataset is especially useful because it gives us a reliable way to classify each country as democratic or not for any given year, while also providing the broader political context needed to understand how and why governments change over time.

The second dataset comes from the World Bank, which tracks global progress on the UN's 17 Sustainable Development Goals, a set of goals aimed at making the world more equitable and sustainable across areas like poverty, education, and gender equality. The 5th goal focuses specifically on gender equality, and one of its key indicators is the percentage of parliamentary seats held by women in each country. This figure is compiled by the World Bank using data from the Inter-Parliamentary Union, an organization that collects official seat counts directly from legislative bodies around the world. It is available annually for most countries from 1997 to 2020, with coverage becoming more consistent in the later years of that range.

**Sources**

[Democracy Data](https://www.christianbjoernskov.com/bjoernskovrodedata/)

[World Bank Data](https://databank.worldbank.org/source/sustainable-development-goals-(sdgs)#)

## What We Want To Know
1. How does the democratic context of a country shape gender gaps in parliamentary representation and executive leadership over time?

## Exploring Gender Gaps in Leadership and Representation
```{r}
#| echo: false
democracy_data <- democracy_data |>
  mutate(
    president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession = monarch_accession_year - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = ifelse(year < 1960, "1950-1959",
       ifelse(year < 1970, "1960-1969",
       ifelse(year < 1980, "1970-1979",
       ifelse(year < 1990, "1980-1989",
       ifelse(year < 2000, "1990-1999",
       ifelse(year < 2010, "2000-2009",
                           "2010-2020")))))))

summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |> 
    filter(!is.na(democracy_data[[group_var]])) |>
    # grouping data by specified grouping variable
    group_by(across(all_of(group_var))) |>
    summarise(n = n(),
              # calculating proportions of female and male presidents
              prop_female_president = mean(
                ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
              prop_male_president = mean(
                ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
              # calculating difference in proportions of male presidents and female presidents (male - female)
              gap_president_prop = prop_male_president - prop_female_president,
              # calculating the average accession ages for female and male presidents
              avg_accession_age_female_president = mean(
                president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
              avg_accession_age_male_president = mean(
                president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
              # calculating difference in average accession ages of male and female presidents (male - female)
              gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
              
              # calculating proportions of female and male monarchs
              prop_female_monarch = mean(
                ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
              prop_male_monarch = mean(
                ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
              # calculating difference in proportions of male monarchs and female monarchs (male - female)
              gap_monarch_prop = prop_male_monarch - prop_female_monarch,
              # calculating the average accession ages for female and male monarchs
              avg_accession_age_female_monarch = mean(
                monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
              avg_accession_age_male_monarch = mean(
                monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
              # calculating difference in average accession ages of male and female monarchs (male - female)
              gap_monarch_accession_age =
                avg_accession_age_male_monarch - avg_accession_age_female_monarch,
              .groups = "drop")
}

plotMe <- summarise_leadership_gender_gap("year")
```

```{r}
#| echo: false
pal <- wes_palette("Moonrise3")
# Table 1: Proportions
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n,
         prop_female_president, prop_male_president, gap_president_prop,
         prop_female_monarch, prop_male_monarch, gap_monarch_prop) |>
  gt() |>
  tab_header(
    title    = md("**Gender Gaps in Executive Leadership by Democracy Status**"),
    subtitle = "Proportions of Female and Male Leaders") |>
  tab_spanner(
    label   = "Presidents",
    columns = c(prop_female_president, prop_male_president, gap_president_prop)) |>
  tab_spanner(
    label   = "Monarchs",
    columns = c(prop_female_monarch, prop_male_monarch, gap_monarch_prop)) |>
  cols_label(is_democracy = "Democracy Status",
    n = "Sample Size", prop_female_president = "Prop. Female",
    prop_male_president = "Prop. Male",
    gap_president_prop  = "Gap (M - F)",
    prop_female_monarch = "Prop. Female",
    prop_male_monarch = "Prop. Male",
    gap_monarch_prop  = "Gap (M - F)") |>
  fmt_percent(columns  = c(prop_female_president, prop_male_president, gap_president_prop, prop_female_monarch, prop_male_monarch, gap_monarch_prop),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_president_prop,
      rows = gap_president_prop > 0)) |>
  tab_style(style = cell_fill(color = "#FFB6C199"),
    locations = cells_body(
      columns = gap_president_prop,
      rows = gap_president_prop < 0)) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_monarch_prop,
      rows    = gap_monarch_prop > 0)) |>
  tab_style(style = cell_fill(color = "#FFB6C199"),
    locations = cells_body(
      columns = gap_monarch_prop,
      rows = gap_monarch_prop < 0)) |>
  tab_footnote(footnote  = "Positive values (shaded blue) indicate males are higher; negative values (shaded pink) indicate females are higher.",
    locations = cells_column_labels(columns = c(gap_president_prop, gap_monarch_prop))) |>
  opt_table_font(font = google_font("Source Serif 4"))

# Table 2: Accession Ages
summarise_leadership_gender_gap("is_democracy") |>
  mutate(is_democracy = ifelse(is_democracy, "Democracy", "Non-Democracy")) |>
  select(is_democracy, n,
         avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age,
         avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age) |>
  gt() |>
  tab_header(title = md("**Gender Gaps in Accession Age by Democracy Status**"),
    subtitle = "Average Accession Ages for Female and Male Leaders") |>
  tab_spanner(label = "Presidents",
    columns = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age)) |>
  tab_spanner(label = "Monarchs",
    columns = c(avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age)) |>
  cols_label(
    is_democracy = "Democracy Status",
    n = "Sample Size",
    avg_accession_age_female_president = "Avg. Age (Female)",
    avg_accession_age_male_president = "Avg. Age (Male)",
    gap_president_accession_age = "Gap (M - F)",
    avg_accession_age_female_monarch  = "Avg. Age (Female)",
    avg_accession_age_male_monarch = "Avg. Age (Male)",
    gap_monarch_accession_age = "Gap (M - F)") |>
  fmt_number(columns  = c(avg_accession_age_female_president, avg_accession_age_male_president, gap_president_accession_age, avg_accession_age_female_monarch, avg_accession_age_male_monarch, gap_monarch_accession_age),
    decimals = 1) |>
  fmt_integer(columns = n) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_spanners()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_column_labels()) |>
  tab_style(style = cell_fill(color = "#D3D3D3"),
    locations = cells_title()) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_president_accession_age,
      rows = gap_president_accession_age > 0)) |>
  tab_style(style = cell_fill(color = "#FFB6C199"),
    locations = cells_body(
      columns = gap_president_accession_age,
      rows  = gap_president_accession_age < 0)) |>
  tab_style(style = cell_fill(color = paste0(pal[1], "99")),
    locations = cells_body(
      columns = gap_monarch_accession_age,
      rows = gap_monarch_accession_age > 0)) |>
  tab_style(style = cell_fill(color = "#FFB6C199"),
    locations = cells_body(
      columns = gap_monarch_accession_age,
      rows = gap_monarch_accession_age < 0)) |>
  tab_footnote(footnote  = "Positive values (shaded blue) indicate males are higher; negative values (shaded pink) indicate females are higher.",
    locations = cells_column_labels(columns = c(gap_president_accession_age, gap_monarch_accession_age))) |>
  opt_table_font(font = google_font("Source Serif 4"))
```

The proportion of female presidents is higher for democratic countries than non-democratic countries. The proportions of male presidents are similar across democratic and non-democratic countries, but the proportion of male presidents for democratic countries is slightly lower. Non-democratic countries' female presidents are, on average, about 13.7 years older than male presidents. For democratic countries, that difference is much lower, at 2.81 years. 

The proportion of male monarchs in non-democratic countries is 84.1 percentage points higher than that of female monarchs. However, in democratic countries, the proportions of male and female monarchs are very similar (47.8% and 52.2%, respectively). Monarchs of both gender are much younger on average than presidents, around 30-35 years old. Non-democratic countries seem to have younger monarchs than democratic countries. 

Next, we created a ribbon plot to highlight the differences in age between average accession ages of presidents and monarchs, by gender. We omitted years before 1980, as there were no female presidents before then, so we did not have any values for the average female presidents' accession age.

```{r}
#| echo: false
#| warning: false
#| fig-alt: This is a ribbon plot, depicting the differences in average accession ages of presidents and monarchs by gender. There are four lines on the graph, one for each gender and leader type combination. The x-axis is years 1980-2020, and the y-axis shows ages 20-70. 

moonrise3 <- wes_palette("Moonrise3")

plotMe %>%
  filter(year >= 1980) %>%
  ggplot(aes(x = year)) +
  geom_ribbon(data = plotMe, aes(x = year,
                                 ymin = avg_accession_age_female_president,
                                 ymax = avg_accession_age_male_president),
              fill = "lightgray", alpha = 0.5) +
  geom_ribbon(data = plotMe,
              mapping = aes(x = year,
                            ymin = avg_accession_age_female_monarch,
                            ymax = avg_accession_age_male_monarch),
              fill = "lightgray", alpha = 0.5) +
  geom_line(aes(y = avg_accession_age_female_president, color = "Female President"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_president,   color = "Male President"),   linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_female_monarch,   color = "Female Monarch"),   linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_monarch,     color = "Male Monarch"),     linewidth = 1.5) +
  scale_color_manual(
    name = "Leader Type",
    values = c(
      "Female President" = moonrise3[3],
      "Male President"   = moonrise3[2],
      "Male Monarch"     = moonrise3[4],
      "Female Monarch"   = moonrise3[1]),
    breaks = c("Female President", "Male President", "Male Monarch", "Female Monarch")) +
  scale_x_continuous(limits = c(1980, 2020)) +
  scale_y_continuous(limits = c(20, 70)) +
  labs(title = "Average Accession Ages of Presidents & Monarchs by Gender",
    x     = "Year",
    y     = "Average Accession Age") +
  theme_minimal()
```

From 1980 to 2020, gender differences in the age at which leaders take office vary between presidents and monarchs. Among presidents, the gap is small and relatively stable, with men only slightly older than women at the time they take office. Both groups also show a gradual increase in age over time, suggesting that gender plays little role in when presidents begin their terms. In contrast, monarchs show a much larger and growing gap. Male monarchs tend to take the throne at older ages, while female monarchs are generally younger, with their ages even decreasing in more recent years. Overall, gender disparities in age at taking office are minimal in elected systems but much more pronounced in monarchical ones, where succession patterns create stronger divides.

## Exploring Women's Representation in National Parliament

```{r}
#| echo: false
#| fig-alt: This is a boxplot depicting the proportion of women in parliament, while faceting on the democracy status of the country. The mean proportion of women holding seats in non-democratic countries is 0.16 and 0.177 for democratic countries. The median proportion of women holding seats in non-democratic countries is 0.14 and 0.158 for democratic countries.

pal <- wes_palette("Moonrise3", n = 2)  

democracy_data %>%
  filter(!is.na(is_democracy)) %>%
  filter(!is.na(prop_in_par)) %>%
  mutate(is_democracy = factor(is_democracy, labels = c("No", "Yes"))) %>%
  ggplot(aes(x = is_democracy, y = prop_in_par)) +
  theme(text = element_text(size = 11, family = "serif")) +
  geom_boxplot(fill = pal,              
    color = "black",          
    outlier.color = "black") +
  stat_summary(fun = mean,
    geom = "crossbar",
    color = "black",
    width = 0.75,
    linetype = "dotted",
    size = .2) +
  stat_summary(fun = mean,
    geom = "text",
    aes(label = paste0("Mean: ", round(after_stat(y), 1),"%")),
    vjust = -1,
    color = "black") +
  stat_summary(fun = median,
    geom = "text",
    aes(label = paste0("Median: ", round(after_stat(y), 1),"%")),
    vjust = 1.5,
    color = "black") +
  labs(title = "Women's Representation in National Parliament by Democracy Status",
    x = "Democracy Status",
    y = "Percentag of National Parliament Seats Held by Women") + 
  theme(plot.title = element_text(size = 5),
    axis.title = element_text(size = 5)) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  theme_minimal()
```

Democracies tend to have slightly higher representation of women in national parliaments compared to non-democracies. The median percentage is higher for democracies (15.8%) than for non-democracies (14%), and the mean is also somewhat higher (17.7% vs. 16%).

However, the distributions are fairly similar overall, with substantial overlap between the two groups’ boxplots. Both democracies and non-democracies show wide variability and several outliers. Non-democratic countries seem to have more outliers, including some at higher values, with two outliers exceeding 60%.

## Fitting a model

We ran a t-test to compare the means of the proportion of national parliament seats held by women between different democracy statuses.

```{r}
#|echo: false
result <- t.test(prop_in_par ~ is_democracy, data = democracy_data)
table_result <- tidy(result)

table_result |>
  select(estimate1, estimate2, statistic, p.value, conf.low, conf.high, parameter) |>
  gt() |>
  tab_header(
    title    = md("**Proportion in Parliament by Democracy Status**"),
    subtitle = "Welch Two-Sample t-Test") |>
  cols_label(estimate1  = md("Mean<br>(Non-Democracy)"),
    estimate2 = md("Mean<br>(Democracy)"),
    statistic = md("*t*-statistic"),
    p.value = md("*p*-value"),
    conf.low = "95% CI Low",
    conf.high = "95% CI High",
    parameter = md("*df*")) |>
  fmt_number(columns  = c(estimate1, estimate2, statistic, conf.low, conf.high),
    decimals = 3) |>
  fmt_number(columns  = parameter,
    decimals = 1) |>
  fmt_scientific(columns  = p.value,
    decimals = 3) |>
  tab_style(style = cell_fill(color = "#fff3cd"),
    locations = cells_body(columns = p.value)) |>
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_body(columns = p.value)) |>
  tab_spanner(label = "Group Means",
    columns = c(estimate1, estimate2)) |>
  tab_spanner(label = "Confidence Interval",
    columns = c(conf.low, conf.high)) |>
  tab_footnote(footnote = "Highlighted cell indicates a statistically significant result.",
    locations = cells_column_labels(columns = p.value)) |>
  opt_stylize(style = 6, color = "blue") |>
  opt_table_font(font = google_font("Source Serif 4"))
```

The t-statistic is -4.591 and the p-value is very small. Also, the confidence interval does not include zero, so we reject the null hypothesis and conclude that there is a statistically significant difference in the mean proportion of seats held by women between democratic and non-democratic countries. This suggests that the true difference in means is not zero.