---
title: "STAT 431 Project 1 Final Deliverable"
author: "Julia Aptekar"
embedded-resources: true
code-fold: true
format: html
---
```{r}
#| echo: false
#install.packages("tidyverse")
#install.packages("wesanderson")
options(pillar.width = Inf)
library(tidyverse)
library(wesanderson)
democracy_data <- read.csv("Combined_Democracy_Data.csv")
```

## Explore the data

The dataset we used was put together to conduct research on political transitions for countries around the world. Its purpose is to collect information on regime characteristics, specifically focusing on how a country's political system changes over time. The dataset consists of 14,768 observations and 43 variables. Each row represents political information for one country on a given year, ranging from 1950 to 2020. The variables provide information on whether a country is democratic, its regime type, and information on elections, leaders, suffrage, power changes, and colonial history.

[Democracy Data](https://www.christianbjoernskov.com/bjoernskovrodedata/)

We also collected data from the World Bank's Sustainable Development Goals database, organized specifically around the UN’s 17 Sustainable Development Goals. The SDGs are separated by topic (e.g., poverty, education, gender equality). It provides country-year panel data (typically yearly) covering economic, social, environmental, and institutional statistics aligned with each goal.

The 5th SDG is "Achieve gender equality and empower all women and girls." 
The proportion of women in national parliaments variable is under the 5th SDG. This variable measures the percentage of seats held by women in a country's national parliament, typically focusing on the lower or single legislative chamber. It is reported annually and serves as an indicator of gender equality in politics.

This proportion is available for most countries from 1997-2020. There are missing values for certain years and countries, but the proportion is more available during the later years of that range.

- [SDG 5 description](https://sdgs.un.org/goals/goal5)
- [Where we found the data](https://databank.worldbank.org/source/sustainable-development-goals-(sdgs)#)
- [Variable description from World Bank](https://databank.worldbank.org/metadataglossary/world-development-indicators/series/SG.GEN.PARL.ZS)

## RQs
1. How does the democratic context of a country shape gender gaps in parliamentary representation and executive leadership over time?

## Exploring Gender Gaps in Leadership and Representation
```{r}
#| echo: false
democracy_data <- democracy_data |>
  mutate(
    president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession = monarch_accession_year - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = ifelse(year < 1960, "1950-1959",
       ifelse(year < 1970, "1960-1969",
       ifelse(year < 1980, "1970-1979",
       ifelse(year < 1990, "1980-1989",
       ifelse(year < 2000, "1990-1999",
       ifelse(year < 2010, "2000-2009",
                           "2010-2020")))))))

summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |> 
    filter(!is.na(democracy_data[[group_var]])) |>
    # grouping data by specified grouping variable
    group_by(across(all_of(group_var))) |>
    summarise(n = n(),
              # calculating proportions of female and male presidents
              prop_female_president = mean(
                ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
              prop_male_president = mean(
                ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
              # calculating difference in proportions of male presidents and female presidents (male - female)
              gap_president_prop = prop_male_president - prop_female_president,
              # calculating the average accession ages for female and male presidents
              avg_accession_age_female_president = mean(
                president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
              avg_accession_age_male_president = mean(
                president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
              # calculating difference in average accession ages of male and female presidents (male - female)
              gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
              
              # calculating proportions of female and male monarchs
              prop_female_monarch = mean(
                ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
              prop_male_monarch = mean(
                ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
              # calculating difference in proportions of male monarchs and female monarchs (male - female)
              gap_monarch_prop = prop_male_monarch - prop_female_monarch,
              # calculating the average accession ages for female and male monarchs
              avg_accession_age_female_monarch = mean(
                monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
              avg_accession_age_male_monarch = mean(
                monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
              # calculating difference in average accession ages of male and female monarchs (male - female)
              gap_monarch_accession_age =
                avg_accession_age_male_monarch - avg_accession_age_female_monarch,
              .groups = "drop")
}

plotMe <- summarise_leadership_gender_gap("year")
summarise_leadership_gender_gap("is_democracy")
```
The proportion of female presidents is higher for democratic countries than non-democratic countries. The proportions of male presidents are similar across democratic and non-democratic countries, but the proportion of male presidents for democratic countries is slightly lower. Non-democratic countries' female presidents are, on average, about 13.7 years older than male presidents. For democratic countries, that difference is much lower, at 2.81 years. 

The proportion of male monarchs in non-democratic countries is 84.1 percentage points higher than that of female monarchs. However, in democratic countries, the proportions of male and female monarchs are very similar (47.8% and 52.2%, respectively). Monarchs of both gender are much younger on average than presidents, around 30-35 years old. Non-democratic countries seem to have younger monarchs than democratic countries. 

Next, we created a ribbon plot to highlight the differences in age between average accession ages of presidents and monarchs, by gender. We omitted years before 1980, as there were no female presidents before then, so we did not have any values for the average female presidents' accession age.

```{r}
#| echo: false
#| fig-alt: This is a ribbon plot, depicting the differences in average accession ages of presidents and monarchs by gender. There are four lines on the graph, one for each gender and leader type combination. The x-axis is years 1980-2020, and the y-axis shows ages 20-70. 

moonrise3 <- wes_palette("Moonrise3")

plotMe %>%
  filter(year >= 1980) %>%
  ggplot(aes(x = year)) +
  geom_ribbon(data = plotMe, aes(x = year,
                                 ymin = avg_accession_age_female_president,
                                 ymax = avg_accession_age_male_president),
              fill = "lightgray", alpha = 0.5) +
  geom_ribbon(data = plotMe,
              mapping = aes(x = year,
                            ymin = avg_accession_age_female_monarch,
                            ymax = avg_accession_age_male_monarch),
              fill = "lightgray", alpha = 0.5) +
  geom_line(aes(y = avg_accession_age_female_president, color = "Female President"), linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_president,   color = "Male President"),   linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_female_monarch,   color = "Female Monarch"),   linewidth = 1.5) +
  geom_line(aes(y = avg_accession_age_male_monarch,     color = "Male Monarch"),     linewidth = 1.5) +
  scale_color_manual(
    name = "Leader Type",
    values = c(
      "Female President" = moonrise3[3],
      "Male President"   = moonrise3[2],
      "Male Monarch"     = moonrise3[4],
      "Female Monarch"   = moonrise3[1]),
    breaks = c("Female President", "Male President", "Male Monarch", "Female Monarch")) +
  scale_x_continuous(limits = c(1980, 2020)) +
  scale_y_continuous(limits = c(20, 70)) +
  labs(title = "Average Accession Ages of Presidents & Monarchs by Gender",
    x     = "Year",
    y     = "Average Accession Age") +
  theme_minimal()
```

From 1980 to 2020, gender differences in the age at which leaders take office vary between presidents and monarchs. Among presidents, the gap is small and relatively stable, with men only slightly older than women at the time they take office. Both groups also show a gradual increase in age over time, suggesting that gender plays little role in when presidents begin their terms. In contrast, monarchs show a much larger and growing gap. Male monarchs tend to take the throne at older ages, while female monarchs are generally younger, with their ages even decreasing in more recent years. Overall, gender disparities in age at taking office are minimal in elected systems but much more pronounced in monarchical ones, where succession patterns create stronger divides.

## Exploring Women's Representation in National Parliament

```{r}
#| echo: false
#| fig-alt: This is a boxplot depicting the proportion of women in parliament, while faceting on the democracy status of the country. The mean proportion of women holding seats in non-democratic countries is 0.16 and 0.177 for democratic countries. The median proportion of women holding seats in non-democratic countries is 0.14 and 0.158 for democratic countries.

pal <- wes_palette("Moonrise3", n = 2)  

democracy_data %>%
  filter(!is.na(is_democracy)) %>%
  filter(!is.na(prop_in_par)) %>%
  mutate(is_democracy = factor(is_democracy, labels = c("No", "Yes"))) %>%
  ggplot(aes(x = is_democracy, y = prop_in_par)) +
  theme(text = element_text(size = 11, family = "serif")) +
  geom_boxplot(fill = pal,              
    color = "black",          
    outlier.color = "black") +
  stat_summary(fun = mean,
    geom = "crossbar",
    color = "black",
    width = 0.75,
    linetype = "dotted",
    size = .2) +
  stat_summary(fun = mean,
    geom = "text",
    aes(label = paste0("Mean: ", round(after_stat(y), 1),"%")),
    vjust = -1,
    color = "black") +
  stat_summary(fun = median,
    geom = "text",
    aes(label = paste0("Median: ", round(after_stat(y), 1),"%")),
    vjust = 1.5,
    color = "black") +
  labs(title = "Women's Representation in National Parliament by Democracy Status",
    x = "Democracy Status",
    y = "National Parliament Seats Held by Women (%)") + 
  theme(plot.title = element_text(size = 5),
    axis.title = element_text(size = 5)) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  theme_minimal()
```

Democracies tend to have slightly higher representation of women in national parliaments compared to non-democracies. The median percentage is higher for democracies (15.8%) than for non-democracies (14%), and the mean is also somewhat higher (17.7% vs. 16%).

However, the distributions are fairly similar overall, with substantial overlap between the two groups’ boxplots. Both democracies and non-democracies show wide variability and several outliers. Non-democratic countries seem to have more outliers, including some at higher values, with two outliers exceeding 60%.