---
title: "STAT 431 Project 1 Checkpoint 2"
author: "Julia Aptekar"
format: html
---

## Context
This is a long data frame with 14,768 observations and 43 variables. 
Each row contains political information for one country during a particular year.
The variables of the data frame provide information on whether a country is democratic, what type of regime the country has, the structure of the political system, along with information on elections, leaders, suffrage, power changes, and colonial history. 
There are variables about whether the leader is a president or a monarch, when their leadership term began, their gender, and their birth year. 
Three variables provide information on timing of regime or leadership changes. 
Lastly, three variables provide information on colonial status and relationships of a country.

## What Cleaning Has Been Done?
First, variables were renamed to all lowercase names with clean_names(). Then, many variable names were renamed to clearer and more descriptive names (for example, dd_regime became regime_category_index). Next, "?" values in the election_month column were converted to NA with na_if(), before the column was split into separate month and year variables with separate_wider_regex(). A new variable, electoral_category, is created using case_match() from electoral_category_index which contained values of 0-3 that measure the level of electoral competition; a 0 means no elections, a 1 means single-party elections, a 2 means non-democratic multi-party elections, and a 3 means democratic elections. Also, str_squish() was used to clean up whitespaces in the election_month variable. Lastly, variables were converted to appropriate data types using across(); year/count columns became numeric and indicator variables (starting with _is or _has) became logical.

## Packages & Data Cleaning
```{r}
library(tidyverse)
options(pillar.width = Inf)
democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

## How have gender gaps in presidential and monarchical leadership evolved across decades, and are these gaps larger in non-democratic countries?
**Modified columns with a custom function**
We created four new variables.

- president_age_at_accession: a president's age at their accession, calculated by subtracting the president's birth year from their accession year
- monarch_age_at_accession: a monarch's age at their accession, calculated by subtracting the monarch's birth year from their accession year
- has_female_leader: an indicator variable indicating whether or not a country had a female president or monarch for that year, a 1 is "yes" and a 0 is "no"
- decade: decade category, split up the years evenly using nested ifelse() statements

Afterwards, we wrote a custom function summarise_leadership_gender_gap() that uses an input grouping variable to group the data before calculating proportions of female and male monarchs and presidents, their average accession ages, and the difference in those proportions and ages.
```{r}
# creating our new variables
democracy_data <- democracy_data |>
  mutate(
    president_age_at_accession = president_accesion_year - president_birthyear,
    monarch_age_at_accession = monarch_accession_year - monarch_birthyear,
    has_female_leader = ifelse(is_female_president | is_female_monarch, 1, 0),
    decade = ifelse(year < 1960, "1950-1959",
       ifelse(year < 1970, "1960-1969",
       ifelse(year < 1980, "1970-1979",
       ifelse(year < 1990, "1980-1989",
       ifelse(year < 2000, "1990-1999",
       ifelse(year < 2010, "2000-2009",
                           "2010-2020")))))))

# writing our custom function
summarise_leadership_gender_gap <- function(group_var) {
  democracy_data |> 
    filter(!is.na(democracy_data[[group_var]])) |>
    # grouping data by specified grouping variable
    group_by(across(all_of(group_var))) |>
    summarise(n = n(),
      # calculating proportions of female and male presidents
      prop_female_president = mean(
        ifelse(!is_monarchy, is_female_president, NA), na.rm = TRUE),
      prop_male_president = mean(
        ifelse(!is_monarchy, !is_female_president, NA), na.rm = TRUE),
      # calculating difference in proportions of male presidents and female presidents (male - female)
      gap_president_prop = prop_male_president - prop_female_president,
      # calculating the average accession ages for female and male presidents
      avg_accession_age_female_president = mean(
        president_age_at_accession[is_presidential & is_female_president], na.rm = TRUE),
      avg_accession_age_male_president = mean(
        president_age_at_accession[is_presidential & !is_female_president], na.rm = TRUE),
      # calculating difference in average accession ages of male and female presidents (male - female)
      gap_president_accession_age = avg_accession_age_male_president - avg_accession_age_female_president,
      
      # calculating proportions of female and male monarchs
      prop_female_monarch = mean(
        ifelse(is_monarchy, is_female_monarch, NA), na.rm = TRUE),
      prop_male_monarch = mean(
        ifelse(is_monarchy, !is_female_monarch, NA), na.rm = TRUE),
      # calculating difference in proportions of male monarchs and female monarchs (male - female)
      gap_monarch_prop = prop_male_monarch - prop_female_monarch,
      # calculating the average accession ages for female and male monarchs
      avg_accession_age_female_monarch = mean(
        monarch_age_at_accession[is_monarchy & is_female_monarch], na.rm = TRUE),
      avg_accession_age_male_monarch = mean(
        monarch_age_at_accession[is_monarchy & !is_female_monarch], na.rm = TRUE),
      # calculating difference in average accession ages of male and female monarchs (male - female)
      gap_monarch_accession_age =
        avg_accession_age_male_monarch - avg_accession_age_female_monarch,
      .groups = "drop")
}
```

We can add any variable from the dataset as a grouping variable. For example, we could input is_democracy to see the different values for democratic vs non-democratic countries or countries with and without free and fair elections.
```{r}
summarise_leadership_gender_gap("is_democracy")
```
The proportion of female presidents is higher for democratic countries than non-democratic countries. The proportions of male presidents are similar across democratic and non-democratic countries, but the proportion of male presidents for democratic countries is slightly lower. Non-democratic countries' female presidents are, on average, about 13.7 years older than male presidents. For democratic countries, that difference is much lower, at 2.81 years. 

The proportion of male monarchs in non-democratic countries is 84.1 percentage points higher than that of female monarchs. However, in democratic countries, the proportions of male and female monarchs are very similar (47.8% and 52.2%, respectively). Monarchs of both gender are much younger on average than presidents, around 30-35 years old. Non-democratic countries seem to have younger monarchs than democratic countries. 

We also want to see these different values by decade.
```{r}
summarise_leadership_gender_gap("decade")
```
**Conclusion**

Over time, the proportions of female presidents has increased to 6.96% of countries having a female president by 2020. The gender gap for presidents has decreased by 13.7 percentage points for female and male presidents. The average accession ages for female and male presidents are similar across all years, typically around 50-55 years old. The largest difference in accession ages occurred in 1970-1979, when the average accession age of female presidents was 43, and the average accession age of male presidents was 48.6.

The proportion of female monarchs was increasing until the 2000-2009 decade, where it slightly dropped from 35.6% to 32.6%. After that, it dropped to 25.3% for the 2010-2020 decade. This is the lowest proportion of female monarchs across all decades; thus the gender gap for monarchs increased from 41.7% in 1950-1959 to 49.5% in 2010-2020. Average accession ages for monarchs are much younger than that of presidents. The typical accession age for a monarch is around 30 years. Female monarchs seem to be slightly younger than male monarchs at accession. The most dramatic difference in accession ages is seen during 2010-2020, where the average male monarch accession age was 44.1, and the average female monarch accession age was 28.6. 